---
title: "CopMI: scales, assumptions, staged fitting, and diagnostics"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{CopMI: scales, assumptions, staged fitting, and diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(CopMI)
```

## NHANES example data

`nhanes_pah` contains six urinary PAH metabolites from the manuscript's 1,330
NHANES 2015-2016 participants. The concentrations are real and the censoring is
artificial, fixed for this teaching example. URXP01 is censored below 346.9 ng/L
(133 values, 10%); URXP03 below 43 ng/L (396 values, 29.77%); URXP04 below
80 ng/L (265 values, 19.92%); and URXP06 below 87 ng/L (529 values, 39.77%).
URXP02 and URXP25 remain observed. These are illustrative thresholds, not laboratory
LODs or a particular evaluation replicate. This example does not establish
normality after logging.

The packaged R object is already on the natural-log scale. Its `ind` matrix uses
**0 = censored, 1 = observed**; `X_cens` stores log(LOD) placeholders in censored
cells. Raw-scale CSV data with NA in censored cells and matching `_ind` columns
are installed under `system.file("extdata", package = "CopMI")`, together with a
variable/threshold table. See `?nhanes_pah`. Neither complete exposure truth nor
data-generation/censoring-generation code is shipped.

Source: CDC/NCHS, [NHANES 2015-2016 PAH_I](https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2015/DataFiles/PAH_I.htm),
linked to DEMO_I, DPQ_I, BMX_I and ALB_CR_I for cohort selection. Original public
files are freely available from NHANES. This derived example is not endorsed by
CDC, HHS or the US Government. Examples illustrate imputation and do not perform
survey-weighted population inference.

## State the modelling assumption explicitly

CopMI separates the supplied data scale from the marginal-distribution choice.
Use `input_scale = "raw"` to log positive raw values internally, or `"log"` for
values already logged. Use `margin_mode = "normal"` to assume normal margins on
the log scale. Use `"select"` (alias `"sd_shift"`) for BIC selection after an
SD-based shift. The method does not run a normality test to choose this setting.

```{r input}
data(nhanes_pah)
dat <- make_lod_data(nhanes_pah$X_cens,
                     nhanes_pah$ind, nhanes_pah$cutoffs)
dat
colMeans(dat$ind == 0L)
```

The indicators define censoring: `1` observed and `0` left-censored. Censored
cells may initially contain `NA`; the constructor replaces them with cutoffs.
The data require one finite cutoff per censored column and at least two
observed values per margin, with variation required except for exponential margins for fitting. An observed value below its
cutoff is rejected. No complete exposure truth is included in the package.

## Run the complete normal-margin workflow

```{r full}
fit <- copula_em_impute(dat, margin_mode = "normal", input_scale = "log",
                         m = 2, max_iter = 100, seed = 42)
summary(fit)
head(CopMI::complete(fit, 1))
```

Inspect `converged`; reaching a chosen iteration limit does not imply
convergence. The flag means that the maximum absolute change in the correlation
matrix fell below `tol`. It does not certify a global likelihood optimum.

## Reproduce the result step by step

```{r stages}
margins <- copmi_fit_margins(dat, margin_mode = "normal", input_scale = "log")
margins$fits[[1]]$parameters
margins$candidate_table
latent <- copmi_transform(margins)
head(latent$Z)
latent$z_lod
model <- copmi_fit_copula(latent, max_iter = 100, seed = 42)
model$Sigma_hat
model$em_change_history
staged <- copmi_impute(model, m = 2)
stopifnot(identical(CopMI::complete(staged), CopMI::complete(fit)))
```

`copmi_fit_copula()` estimates the model without drawing completed matrices.
`copmi_impute()` can reuse the fitted model without repeating marginal fitting or
EM. Its default seed continues the stored fitting RNG state; a different integer
seed requests a new run of the conditional sampler. Explicitly seeded calls
restore the caller's random-number state.

## Select heterogeneous margins

The original default candidate set contains 11 families. Here a short candidate
list makes the decision table easier to inspect.

```{r selection}
selected <- copmi_fit_margins(dat, margin_mode = "select",
                              margin_candidates = c("norm", "logis", "lnorm"))
selected$family_selected
selected$candidate_table
selected$shift$anchor
```

For each variable the shift is `anchor = cutoff - shift_k * observed_sd`,
with the minimum observed value as the reference if the cutoff is absent. The
working values equal logged input minus the anchor. Marginal parameters in this
mode refer to that shifted scale, not the original exposure scale. Positive-support
selected families imply values above the anchor. Selected normal and logistic
families retain real support; their values are not forcibly made positive.

To inspect one margin independently:

```{r single}
single <- copmi_fit_margin(dat$X_cens[, 1], dat$ind[, 1], dat$cutoffs[1],
                           candidates = "norm")
single$parameters
single$bic
```

This lower-level single-column function fits exactly the scale provided to it;
it does not perform logging or shifting. For automatic scale handling use
`copmi_fit_margins()` or `copula_em_impute()`.

## L-BFGS-B first, Nelder-Mead as Plan B

For each family-specific start, the default order is
`optim_methods = c("L-BFGS-B", "Nelder-Mead")`. A successful finite converged
primary fit does not run the backup. An error, nonconvergence, or invalid fit
activates Nelder-Mead on the same left-censored likelihood and
log-transformed positive parameters, without empirical upper bounds. Candidate selection only uses successfully fitted candidates.

```{r optimizer}
single$optimization
backup_only <- copmi_fit_margin(dat$X_cens[, 1], dat$ind[, 1], dat$cutoffs[1],
                                candidates = "norm", optim_methods = "Nelder-Mead")
backup_only$optimizer
backup_only$parameters
```

The second example explicitly requests the backup optimizer to illustrate its
output; it is not evidence of a naturally occurring primary failure. Automated
tests also force primary failures and verify that the retry really occurs.

## Raw-scale input and output

```{r raw}
table <- read.csv(system.file("extdata", "nhanes_pah_censored.csv", package = "CopMI"))
meta <- read.csv(system.file("extdata", "nhanes_pah_metadata.csv", package = "CopMI"))
X <- as.matrix(table[meta$variable])
ind <- as.matrix(table[paste0(meta$variable, "_ind")])
dimnames(X) <- dimnames(ind) <- list(as.character(table$SEQN), meta$variable)
raw <- make_lod_data(X, ind, meta$cutoff_raw, scale = "raw")
raw_fit <- copula_em_impute(raw, input_scale = "raw", margin_mode = "normal",
                            m = 1, max_iter = 100, seed = 42)
raw_completed <- CopMI::complete(raw_fit, 1)
head(raw_completed)
stopifnot(identical(raw_completed[raw$ind == 1L], raw$X_cens[raw$ind == 1L]))
cutoff_matrix <- matrix(raw$cutoffs, nrow(raw_completed), ncol(raw_completed),
                        byrow = TRUE)
stopifnot(all(raw_completed[raw$ind == 0L] < cutoff_matrix[raw$ind == 0L]))
```

Never log the data without also logging its cutoffs. Do not request `"raw"` for
values already logged. CopMI uses natural log, so already transformed inputs
should use the same convention when the intended assumption is log-normality.

## Extract data and inspect diagnostics

```{r extract}
length(CopMI::complete(fit, "all"))
dim(CopMI::complete(fit, "mean"))
head(CopMI::complete(fit, "long"))
dx <- diagnostics(fit)
dx$converged
dx$optimization
dx$init_diagnostics$pair_results
dx$fallback_records
```

The fallback table distinguishes optimization retries, candidate failures,
initial-pair optimizer retries and initial-matrix positive-definite projection.
E-step, sampling, and inverse-transformation failures stop with an error; they
do not produce approximate or fixed-value replacements. Counts represent events;
multiple events can concern the same sample or variable. Empty tables mean no
events of that type were recorded.

## Unconverged models

By default, imputation stops if EM has not converged. Refit with a larger
`max_iter` and inspect the iteration history. A deliberately exploratory run
can use `allow_unconverged = TRUE`; its convergence flag remains false.

```{r exploratory}
initial <- copmi_fit_copula(latent, max_iter = 0, seed = 42)
exploratory <- copmi_impute(initial, m = 1, allow_unconverged = TRUE)
exploratory$converged
```

## Interpretation and development

The current implementation generates conditional imputations at fixed estimated
margin parameters and correlation. It does not draw parameter uncertainty or
family-selection uncertainty. The Gibbs sampler starts inside the truncation region; assess
sampler mixing, sensitivity, and inferential performance for
your application. `complete(fit, "mean")` is a descriptive average, not an
analysis that propagates between-imputation uncertainty. Downstream model
fitting and pooling are outside this package's current scope.

Run `?copula_em_impute` or `?copmi_fit_margin` for complete contracts. All public
functions include executable examples. The source has separate validation,
marginal-fitting, transformation, EM, sampling, and extraction modules. Run
`roxygen2::roxygenise()`, `testthat::test_local()`, `R CMD build`, and
`R CMD check --as-cran` when changing it.
